---
title: "Dependency Management with renv"
format: gfm
editor: visual
---

This notebook contains the code underpinning the Dependency Management with renv in R thread.

Follow me on Twitter [@neilgcurrie](twitter.com/neilgcurrie) and visit [torchdata.io](https://www.torchdata.io).

You can find links to all my threads [here](https://github.com/neilcuz/threads).

## Using renv

To use `renv` we first need to install it.

```{r}
#| eval: false
install.packages("renv")
```

Now we initialise `renv` within the project.

```{r}
#| eval: false
renv::init()
```

Install, update and use packages as normal.

```{r}
#| eval: false
install.packages("dplyr")
library(dplyr)
```

To take a snapshot of the current project state call `snapshot`. If for some reason you run into problems you can call the `restore` function to revert to the previous lock file.

```{r}
#| eval: false
renv::snapshot()
renv::restore()
```

For a new user to recreate the project library in the same state you had it share on GitHub and it should rebuild automatically when they open the project for the first time.

