---
title: "Pivoting with tidyr"
format: gfm
editor: visual
---

This notebook contains the code underpinning the *Pivoting with* tidyr Twitter thread.

-   For online courses, training and consultancy visit my website [shoogle.co](shoogle.co)

-   Subscribe to my YouTube channel for videos on R and data science @shoogle [youtube.com/@shoogle](www.youtube.com/@shoogle)

-   Follow me on Twitter for weekly threads on data and coding @neilgcurrie [twitter.com/neilgcurrie](www.twitter.com/neilgcurrie)

You can find links to all my threads [here](https://github.com/neilcuz/threads).

## Setup

Check's if tidyr is installed, if not it will install and then load.

```{r}
#| output: false
if (!require(tidyr)) install.packages("tidyr")
library(tidyr)
```
## pivot_longer

Pivot data into tidy or long format from messy or wide format.

```{r}
pop_millions_messy <- tibble(country = c("Argentina", "Albania", "Australia"),
                             `2005` = c(39.18, 3.03, 20.2),
                             `2010` = c(41.34, 2.99, 21.5),
                             `2015` = c(43.43, 3.03, 22.8))

pop_millions_tidy <-  pivot_longer(pop_millions_messy, 
                                   cols = c(`2005`, `2010`, `2015`), 
                                   names_to = "year",
                                   values_to = "population")

print(pop_millions_messy)
print(pop_millions_tidy)
```
## pivot_wider

Pivot data into messy or wide format from tidy or long format.

```{r}
pop_millions_messy <- pivot_wider(pop_millions_tidy, names_from = year,
                                  values_from = population)

print(pop_millions_messy)
```

