---
title: "Writing Functions"
format: gfm
editor: visual
---

This notebook contains the code underpinning the Writing Functions thread.

Follow me on Twitter [\@neilgcurrie](twitter.com/neilgcurrie) and visit [torchdata.io](https://www.torchdata.io).

You can find links to all my threads [here](https://github.com/neilcuz/threads).

## Setup

Install the packages if you need to.

```{r}
#| eval: false
install.packages("dplyr")
install.packages("ggplot2")
```

And load them.

```{r}
#| output: false
library(dplyr)
library(ggplot2)
```


## When to write functions

The rule of three.

```{r}
# Bad

sample1 <- sample(c(TRUE, FALSE), size = 10, replace = TRUE)
names(sample1) <- paste0("pos", 1:10)
sample1_true <- sample1[sample1 == TRUE]

sample2 <- sample(c(TRUE, FALSE), size = 15, replace = TRUE)
names(sample2) <- paste0("pos", 1:15)
sample2_true <- sample2[sample2 == TRUE]

sample3 <- sample(c(TRUE, FALSE), size = 20, replace = TRUE)
names(sample3) <- paste0("pos", 1:20)
sample3_true <- sample3[sample3 == TRUE]

# Better

sampler <- function (size) {
  
  my_sample <- sample(c(TRUE, FALSE), size, replace = TRUE)
  names(my_sample) <- paste0("pos", 1:size)

  return(my_sample[my_sample == TRUE])
  
}

sample1 <- sampler(10)
sample2 <- sampler(15)
sampler3 <- sampler(20)
```

## Defining functions

```{r}

# If you cast your mind back to school this is Pythagoras' Theorem. 
# This function calculates side c (the hypotenuse) given sides a and b of a triangle

# The function name is calculate_c, side_a and side_b are arguments

calculate_c <- function (side_a, side_b) {
  
  return(sqrt(side_a ^ 2 + side_b ^ 2))
  
}

# We could extend the function and give the user control of if they want c
# squared or not. Here we have given squared a default value of FALSE.

calculate_c <- function (side_a, side_b, squared = FALSE) {
  
  side_c <- side_a ^ 2 + side_b ^ 2

  if (squared == TRUE) {
    
    return(side_c)
    
  } else {
    
    return(sqrt(side_c))
    
  }
}


```

## Tidy evaluation

This won't work

```{r}
dummy <- tibble(id = 1:20, i = runif(20), j = runif(20), k = runif(20))

transform_data1 <- function(x, min_value, filter_by, sort_by) {
  
  x |> 
    filter(filter_by >= min_value) |> 
    arrange(sort_by) 
  
}

```

```{r}
#| eval: false
transform_data1(dummy, 0.5, j, k) # error!
```
This will work. And an example of using :=

```{r}
# Will work with the curly brackets

transform_data2 <- function(x, min_value, filter_by, sort_by) {
  
  x |> 
    filter({{filter_by}} >= min_value) |> 
    arrange({{sort_by}}) 
  
}

transform_data2(dummy, 0.5, j, k)

# Using := to modify the left hand side or name

transform_data3 <- function(x, min_value, filter_by, sort_by, new_name) {
  
  x |> 
    filter({{filter_by}} >= min_value) |> 
    arrange({{sort_by}}) |> 
    mutate({{new_name}} := {{filter_by}} * {{sort_by}})
    
  
}

transform_data3(dummy, 0.5, i, k, "l")
```

## Error handling for arguments

```{r}

calculate_c <- function (side_a, side_b, squared = FALSE) {
  
  if (side_a <= 0) {
    
    stop ("side_a must be positive")
    
  }
  
  if (side_b <= 0) {
    
    stop ("side_b must be negative")
    
  }
  
  if (length(side_a) != length(side_b)) {
    
    stop ("side_a and side_by must have the same length")
    
  }
  
  if (!is.logical(squared)) {
    
    stop ("squared must be TRUE or FALSE")
    
  }
  
  side_c <- side_a ^ 2 + side_b ^ 2

  if (squared == TRUE) {
    
    return(side_c)
    
  } else {
    
    return(sqrt(side_c))
    
  }
}

```

## Using ellipsis

```{r}
# Maybe you want to use some of the aesthetics for tweaking geom_point.
# Maybe you arent sure which ones, all you know is you will be doing a bunch
# of scatter plots of i and j.

ggplot(dummy, aes(x = i, y= j)) +
  geom_point(size = 5, shape = 14) +
  labs(title = "My nice plot")

plotter <- function (plot_data, ..., title = "Placeholder title") {
  
  ggplot(plot_data, aes(x = i, y = j)) +
    geom_point(...) +
    labs(title = title)
  
}

plotter(dummy, alpha = 0.5, shape = 10, stroke = 10)

```

## Functions written for their side effects

Let's jut use the plotter function from the last example. Return the first 
argument unmodified and invisibly.

```{r}
plotter2 <- function (plot_data, ..., title = "Placeholder title") {
  
  ggplot(plot_data, aes(x = i, y = j)) +
    geom_point(...) +
    labs(title = title)
  
  return(invisible(plot_data))
  
}

p <- plotter2(dummy, size = 3, shape = 4, stroke = 5)

print(p)

```



