---
title: "Data Wrangling with Arrow in R"
format: gfm
editor: visual
---

This is the code underpinning the thread Data Wrangling with Arrow in R which 
you can find [here]().


### Setup

This installs and loads the required libraries. I have written in this way so 
you can run it without modifying or writing manual installation code.

```{r}
#| results = FALSE, warning = FALSE, message = FALSE
# Install packages (if not already installed)

packages_to_load <- c("arrow", "dplyr", "glue", "here", "readr")
installed <- packages_to_load %in% installed.packages()[,1]
packages_to_install <- packages_to_load[!installed]

install.packages(packages_to_install)

# Load packages

for (package in packages_to_load) library(package, character.only = TRUE)
```

### NY Taxis using readr::read_csv

This code shows the standard approach to reading in some data and manipulating.
This is done with the commonly used functions `readr::read_csv`. As an aside you
will find `vroom::vroom` and `data.table::fread` are faster.

```{r}
file_taxi <- glue("{here::here()}/data/taxi_trip_data.csv")

taxi_standard <- file_taxi |> 
  read_csv(show_col_types = FALSE) |> 
  select(trip_distance, passenger_count, total_amount, pickup_location_id) |> 
  filter(trip_distance >= 5, passenger_count > 1) |> 
  mutate(total_amount = as.numeric(total_amount),
         total_amount_per_head = total_amount / passenger_count) |> 
  group_by(pickup_location_id) |> 
  summarise(mean_amount = mean(total_amount_per_head, na.rm = TRUE))

print(taxi_standard)
```

### NY Taxis using arrow::read_csv_arrow

This code shows the approach for reading in the same data and manipulating in 
the same way, but this time we use the arrow package.


```{r}
taxi_arrow <- file_taxi |> 
  read_csv_arrow() |> 
  select(trip_distance, passenger_count, total_amount, pickup_location_id) |> 
  filter(trip_distance >= 5, passenger_count > 1) |> 
  mutate(total_amount = as.numeric(total_amount),
         total_amount_per_head = total_amount / passenger_count) |> 
  group_by(pickup_location_id) |> 
  summarise(mean_amount = mean(total_amount_per_head, na.rm = TRUE)) |> 
  collect()

print(taxi_arrow)
```

### NIDS using arrow::read_csv_arrow

This code using arrow for reading a larger dataset, one which wouldn't work on
my machine with a standard approach.

```{r}
file_nids <- glue("{here::here()}/data/NF-UQ-NIDS-v2.csv")

nids <- file_nids |>
  read_csv_arrow() |>
  select(IN_PKTS, IN_BYTES) |>
  mutate(IN_PKTS = as.numeric(IN_PKTS),
         IN_BYTES = as.numeric(IN_BYTES)) |>
  filter(IN_PKTS > 3, IN_BYTES > 400) |>
  mutate(bytes_per_packet = IN_BYTES / IN_PKTS) |>
  filter(bytes_per_packet > 200) |>
  collect()

print(nids)
```
