---
title: "How to use R with SQLlite"
output:
  github_document:
    df_print: paged
editor_options:
  markdown:
    wrap: 80
---

This is a walkthrough of how you can use SQLite with R from my twitter [\@neilgcurrie](https://twitter.com/neilgcurrie).

Often in R we read in files of mixed formats. This gets messy fast and when the
data is too big R will lag or fall over. Enter the database. SQLite is software
used to interact with databases and is one of the most popular engines around.
We can work with SQLite directly from R.

Let's create a database. We need the DBI and RSQLite packages to interact with
our database. The dbConnect function creates a connection to a database when you
supply a file name. When no database exists it creates a blank one. This one is
created in my working directory.

```{r}
library(DBI)     
library(RSQLite)

filename <- glue::glue("{here::here()}/football-data.db")

con <- DBI::dbConnect(RSQLite::SQLite(), filename)

DBI::dbListTables(con)
```

Now we add some data to our database. In this example I have taken football
(soccer) results data from football-data.co.uk for the Premier League and La
Liga. We use the dbWriteTable function to create new tables with specified names
and write the data all in one step.

```{r}
library(readr)

# Create a table

epl_2122 <- readr::read_csv("https://www.football-data.co.uk/mmz4281/2122/E0.csv", 
                            show_col_types = FALSE)

DBI::dbWriteTable(con, name = "england_premier_league", value = epl_2122)

dbListTables(con) 

# You can look at a tables fields

DBI::dbListFields(con, name = "england_premier_league")

# Create another table

ll_2122 <- read_csv("https://www.football-data.co.uk/mmz4281/2122/SP1.csv", 
                    show_col_types = FALSE)

dbWriteTable(con, name = "spain_la_liga", value = ll_2122)

dbListTables(con)
```

Imagine we want to add an extra season of Premier League data to our table
england_premier_league. Again, we use the dbWriteTable function but now we use
the option append = TRUE.

```{r}
epl_2021 <- readr::read_csv("https://www.football-data.co.uk/mmz4281/2021/E0.csv",
                            show_col_types = FALSE)

dbWriteTable(con, name = "england_premier_league", value = epl_2021, 
             append = TRUE)
```

Now we have some data in there let's query the database. We can send a SQL query
from R to the database and pull the queried data back into R. We type SQL code
as a normal string and pass it to the dbGetQuery function.

```{r}
query <- "SELECT Date, HomeTeam, AwayTeam, FTHG, FTAG
          FROM   england_premier_league"

football_data1 <- DBI::dbGetQuery(con, statement = query)

print(head(football_data1))

# Say you only wanted games with more than 2.5 goals scored

query <- "SELECT Date, HomeTeam, AwayTeam, FTHG, FTAG
          FROM   england_premier_league
          WHERE  FTHG + FTAG > 2.5"

football_data2 <- DBI::dbGetQuery(con, statement = query)

print(head(football_data2))
```

This is really powerful. This data could be far too big for R to handle all at
once - no problem with this approach. You would only need the memory to handle
the queried, and hopefully smaller, dataset.

What if you don't know SQL? You can use the dbplyr package with all the great
syntax you are familiar with from dplyr! R will turn your R code into SQL code
and send it to the database without needing to know any SQL.

```{r}
library(dplyr, quietly = TRUE, warn.conflicts = FALSE)
library(dbplyr, quietly = TRUE, warn.conflicts = FALSE)

football_data3 <- con |>
  dplyr::tbl("england_premier_league") |>
  dplyr::select(Date, HomeTeam, AwayTeam, FTHG, FTAG) |>
  dplyr::filter(FTHG + FTAG > 2.5)

print(head(football_data3))
```

footballdata3 isn't quite the same as a normal tibble/data.frame. That's because
R hasn't pulled the data in yet. This keeps things fast. To pull it in we use
the collect function.

```{r}
football_data3 <- dplyr::collect(football_data3)
print(football_data3)

football_data2 |> 
  dplyr::as_tibble() |> 
  identical(football_data3) # exactly the same

# remember to close your connection to the database

dbDisconnect(con) 
```

Thanks for reading, if you liked this then follow me on twitter for mainly R and 
data science stuff (though I make no promises) [\@neilgcurrie](https://twitter.com/neilgcurrie).
